---
title: "Evolang_similarity"
author: "Šárka Kadavá"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intro

This is an R Markdown for similarity analysis of sound signal.

First we setup our folders

```{r folders, echo=FALSE}

#get current drive
curfolder   <- dirname(getwd())
project_folder <- paste0(curfolder, "/sound_similarity/")
dataset_folder <- paste0(project_folder, "/datasets/")
audio_path  <- paste0(project_folder, "segments/")
audio_files <- list.files(audio_path,
                         pattern = "*.wav",
                         recursive = TRUE,
                         all.files = FALSE,
                         full.names = TRUE)
audio_orig <- paste0(project_folder, "data/")

audio_detailed <- list.files(audio_orig,
                         pattern = "*.wav",
                         recursive = TRUE,
                         all.files = FALSE,
                         full.names = TRUE)

print(curfolder)
print(audio_path)
```
```{r duration}

# Replace 'your_file.txt' with the path to your .txt file
txt_path <- paste0(project_folder, "duration_log.txt")

# Use read.table if the file is tab-separated or has a different delimiter
# For comma-separated files, you can use read.csv
# Set the 'sep' argument accordingly to specify the delimiter
# In this example, I assume the file is tab-separated
data_df <- read.table(txt_path, header = FALSE, sep = '\t')

# Print the first few rows of the data frame to check the import
head(data_df)

# What is the minimum duration
min(data_df$V5) #18 ms
mean(data_df$V5) # 354 ms


```



## Extracting features from sound

We use package 'soundgen' to extract a package of features for sound signal
More info here: https://www.rdocumentation.org/packages/soundgen/versions/2.6.0 or here: https://cogsci.se/soundgen/acoustic_analysis.html


Now we extract summary of features in segments
```{r soundgen, segments}

library(soundgen)
#library(tuneR)

# Create an empty data frame to store the features
feature_df <- data.frame()

# Create a function to extract summary features from the analyze output
extract_summary_features <- function(file, analyze_result) {
  summary_features <- analyze_result$summary
  # Add the filename as the first column
  summary_features <- cbind(File = file, summary_features)
  return(summary_features)
}

# Loop over the .wav files in the directory and extract summary features
for (file in audio_files) {
  # Check if the file name ends with '_p ', and skip processing if it does
  if (endsWith(file, "_p .wav")) {
    print(paste("Skipping file:", file))
    next  # Skip this file and continue with the next iteration
  }
  print(paste("Analyzing file:", file))
  #audio <- readWave(file)
  roughSet <- list(windowLength=10, step = 3, amRes = 10)
  features <- analyze(file, windowLength = 10, step = 5, roughness = roughSet, pitchCeiling=1000, cutFreq=500)
  # sampling rate needed only if x is numeric vector
  # windowLength 10 ms, the minimum dur of a segment is 18 ms
  
  # Extract and format summary features
  summary_features_df <- extract_summary_features(file, features)
  
  # Append the summary features for this file to the main data frame
  feature_df <- rbind(feature_df, summary_features_df)
}

# Reset row names and remove them
row.names(feature_df) <- NULL

# Print the data frame with features
print(feature_df)

# Create the full file path including the folder and file name
output_file <- paste0(dataset_folder, "/segments_soundgen_summary.csv")

# Save the data frame as a CSV file
write.csv(feature_df, file = output_file, row.names = FALSE)
```

```{r data wrangling}

library(tidyr)
library(dplyr)

data <- feature_df

# Split the 'file' column into separate columns
data <- data %>%
  separate(file, into = c("participant", "trial", "condition", "wordCount", "word", "segmentCount"), sep = "_")

# Remove '.wav' from the 'segmentnumber' column
data$segmentCount <- gsub("\\ .wav", "", data$segmentCount)

# Find columns where the only unique value is NA
na_only_unique_cols <- sapply(data, function(x) length(unique(x, na.rm = TRUE)) == 1 && all(is.na(unique(x, na.rm = TRUE))))

# Extract the column names
cols_to_delete <- names(na_only_unique_cols[na_only_unique_cols])

# Remove the identified columns from the DataFrame
data <- data %>%
  select(-all_of(cols_to_delete))

# Columns to remove because we are not interested in those
additional_cols_to_remove <- c('flux_mean', 'flux_median', 'flux_sd', 'fmDep_mean', 'fmDep_median', 'fmDep_sd', 'fmFreq_mean', 'fmFreq_median', 'fmFreq_sd', 'harmEnergy_mean', 'harmEnergy_median', 'harmEnergy_sd', 'harmHeight_mean', 'harmHeight_median', 'harmHeight_sd')

# Remove the identified columns from the DataFrame
data <- data %>%
  select(-all_of(additional_cols_to_remove))


```

```{r category info}

# Define the mapping between concepts and categories
concept_category_mapping <- data.frame(
  concept = c("no", "happy", "sad", "bad", "good", "angry", "disgusted", "dog", "cat", "bird", "fish", "fly", "old", "spoon", "egg", "ash", "stone-rock", "smoke", "maybe", "not", "scared"),
  category = c("other", "emotion/valence", "emotion/valence", "emotion/valence", "emotion/valence", "emotion/valence", "emotion/valence", "animal", "animal", "animal", "animal", "animal", "time", "thing/object", "thing/object", "thing/object", "thing/object", "thing/object", "abstract/logical", "abstract/logical", "emotion/valence"),
  stringsAsFactors = FALSE
)

# Merge the mapping with your original DataFrame 'df' based on the 'concept' column
data <- merge(data, concept_category_mapping, by.x = "word", by.y = "concept", all.x = TRUE)

# Remove the 'File' column
data <- data %>%
  select(-File)


```



```{r umap}

library(umap) #umap tools
library(ggplot2) #plotting
library(plotly) #for some interactivity in your plots


# what are the features
feats <- data[,7:ncol(data)]
# delete the category
feats <- feats %>%
  select(-category)

feats_scaled <- scale(feats) #normalize the features
feats_scaled[is.na(feats_scaled)] <- 0 #remove all
umap_result <- umap(feats_scaled, n_components = 2) #perform umap

# add umap coordinates to the feature data
umap_df <- data.frame(umap1 = umap_result$layout[, 1], umap2 = umap_result$layout[, 2], word = data$word, participant = data$participant, category = data$category, segment = data$segmentCount, trial = data$trial)

library(viridis)
library(ggforce)

# Generate a list of 50 distinct and non-repeating colors from the 'viridis' palette
distinguishable_colors <- viridis_pal(option = "D")(50)

# plot
color_palette <- c("red", "white", "blue", "pink", "orange")

a <- ggplot(umap_df, aes(x = umap1, y = umap2, color = trial, text = participant)) +
  geom_point(alpha=0.7) +
  geom_line(aes(group = trial), color = "gray", alpha = 0.5) +  # Add this line
  labs(title = "UMAP Plot of Vocal Segments", x = "UMAP Dimension 1", y = "UMAP Dimension 2") +scale_color_manual(name = "Category", values = distinguishable_colors)+ theme_dark()
ggplotly(a)



```


Let's take a look on the inter-segment distance. Some of the words' segments might travel through space more than others

```{r intersegment distance}


umap_df$trial <- as.factor(umap_df$trial)
umap_df$participant <- as.factor(umap_df$participant)

# Function to calculate Euclidean distance between two points
euclidean_distance <- function(x1, y1, x2, y2) {
  sqrt((x1 - x2)^2 + (y1 - y2)^2)
}

# Calculate pairwise Euclidean distances within each trial
avg_distances <- umap_df %>%
  group_by(participant, trial) %>%
  arrange(trial, segment) %>%
  mutate(
    next_x = lead(umap1),
    next_y = lead(umap2)
  ) %>%
  filter(!is.na(next_x) & !is.na(next_y)) %>%
  summarize(Avg_Distance = mean(euclidean_distance(umap1, umap2, next_x, next_y)))

# merge the original df with avg distance

umap_df <- merge(umap_df, avg_distances, by.x = c("trial", "participant"), by.y = c("trial", "participant"))

# what is the distribution
hist(avg_distances$Avg_Distance)

# plot the avg distance between segments per category
ggplot(umap_df, aes(x = category, y = Avg_Distance)) +
  geom_point(size = 3, color = "red") +
  geom_boxplot(width = 0.2, fill = "orange", alpha = 0.5) +  # Add boxplot
  labs(title = "Avg Distance Between Segments Across Categories",
       x = "Category", y = "Average Distance")

```
Do the same, but across words
```{r eucl distance between words}

# Calculate pairwise Euclidean distances for each word
avg_distances_word <- umap_df %>%
  group_by(word) %>%
  mutate(
    next_x = lead(umap1),
    next_y = lead(umap2)
  ) %>%
  filter(!is.na(next_x) & !is.na(next_y)) %>%
  summarize(Avg_Distance_word = mean(euclidean_distance(umap1, umap2, next_x, next_y)))

# merge with main
similarity <- merge(umap_df, avg_distances_word, by.x = "word", by.y = "word")

# plot
ggplot(similarity, aes(x = category, y = Avg_Distance_word)) +
  geom_point(size = 3, color = "red") +
  geom_boxplot(width = 0.2, fill = "orange", alpha = 0.5) +  # Add boxplot
  labs(title = "Avg Distance Between Words Across Categories",
       x = "Category", y = "Average Distance")


```
Alternative: calculate mean of euclDistance between segments per word
```{r}

# Alternative: calculate mean of euclDistance between segments per word
distance_avg_segments_avg_per_word <- umap_df %>%
  group_by(word) %>%
  summarize(mean_euclDist = mean(Avg_Distance))

distance_avg <- merge(umap_df, distance_avg_segments_avg_per_word, by.x = "word", by.y = "word")

ggplot(distance_avg, aes(x = category, y = mean_euclDist)) +
  geom_point(size = 3, color = "red") +
  geom_boxplot(width = 0.2, fill = "orange", alpha = 0.5) +  # Add boxplot
  labs(title = "Avg Distance Between Segments per Word Across Categories",
       x = "Category", y = "Average Distance")


```


Now we extract features as timeseries from the events
```{r soundgen, events}


library(soundgen)
#library(tuneR)


# Create an empty named list to store the detailed features
named_detailed_features_list <- list()


# Loop over the .wav files in the directory and extract summary features
for (file in audio_detailed) {
  print(paste("Analyzing file:", file))

  # Analyze the audio to get detailed features
  detailed_features <- analyze(file, samplingRate = 44100)$detailed
  
    # Extract the file name without the path
  file_name <- basename(file)
  
  # Add the detailed features to the named list with the file name as the key
  named_detailed_features_list[[file_name]] <- detailed_features
}




# Combine the list of data frames into one large data frame
event_features <- do.call(rbind, detailed_features_list)

# Reset row names and remove them
row.names(combined_dataframe) <- NULL

# Create the full file path including the folder and file name
#output_file <- paste0(dataset_folder, "/segments_soundgen_summary.csv")

# Save the data frame as a CSV file
#write.csv(feature_df, file = output_file, row.names = FALSE)

```




